Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-09-15 pot/LaTeX/skewnormal.tex
updates to manuscript

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
We assume the data can be modeled using a skew-$t$ distribution.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N_{(0, \infty)}(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.3), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

@@

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.2.

@@

\section{Computation}\label{s:comp}
The MCMC for this model is fairly straightforward.
First, we impute values below the threshold.
Then, we update $\Theta$ using random walk MH or Gibbs sampling when appropriate.
Finally, we make spatial precictions using conditional multivariate normal results and the fact that the distribution of $Y_t(\bs) \mid \Theta, z_{tl}$ is the usual multivariate normal distribution with a \Matern spatial covariance structure.

We can use Gibbs sampling to update $Y_t(\bs)$ for censored observations that are below the threshold $T$.
After conditioning on $\alpha$, $z_t(\bs)$ and non-censored observations, $Y_t(\bs)$ has truncated normal full conditionals.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}(\mu(\bs), \Sigma)$.
After imputing the censored observations, we update the model parameters.
To update the model parameters, we use standard Gibbs updates for parameters when possible.
In the case Gibbs sampling is not possible, parameters are updated using a random-walk Metropolis Hastings algorithm.
See Appendix A.1 for details regarding Gibbs sampling and Metropolis update rules.
The final step of the computation is to use Bayesian Kriging to generate a predictive distribution for $Y_t(\bs^*)$ at prediction location $(\bs^*)$.
This step is similar to the imputation for censored observations except that the full conditionals are no longer truncated at $T$.

@@

$M = 50$ data sets are generated for each design.
For each data set we fit the data using
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=1$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=1$ knots, $T=q(0.80)$
  \item skew-$t$ marginal, $K=5$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=5$ knots, $T=q(0.80)$
\end{enumerate}
where $q(0.80)$ is the 80th sample quantile of the data.
The design matrix $\bX$ includes an the intercept with a prior of $\beta \sim \text{N}(0, 10)$.
The spatial covariance parameters have priors $\log(\nu) \sim \text{N}(-1.2, 1)$, $\gamma \sim \text{U}(0, 1)$, $\log(\rho) \sim \text{N}(-2, 1)$.
The skewness parameter has prior $\alpha \sim \text{N}(0, 2)$.
The residual variance terms have priors $\sigma^2_t(\bs) \sim \text{IG}(0.1, 0.1)$.

@@

\section*{Appendix A.1: Posterior distributions}
\input{zpost.tex}
\input{betapost.tex}
\input{sigpost.tex}
\input{alphapost.tex}

\section*{Appendix A.2: Proof that $\lim_{h \rightarrow \infty} \pi(h) = 0$}
\input{proof_dist.tex}

\section*{Appendix A.3: Half-normal distribution}
Let $u = |z|$ where $Z \sim N(\mu, \sigma^2)$.
Specifically, we consider the case where $\mu = 0$. Then $U$ follows a half-normal distribution which we denote as $U \sim HN(0, 1)$, and the density is given by
\begin{align}
  f_U(u) = \frac{ \sqrt{2} }{ \sqrt{\pi \sigma^2} } \exp \left( - \frac{ u^2 }{ 2 \sigma^2 } \right) I(u > 0)
\end{align}
When $\mu = 0$, the half-normal distribution is also equivalent to a $N_{(0, \infty)}(0, \sigma^2)$ where $N_{(a, b)}(\mu, \sigma^2)$ represents a normal distribution with mean $\mu$ and standard deviation $\sigma$ that has been truncated below at $a$ and above at $b$.

>>>>

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
We assume the data can be modeled using a skew-$t$ distribution.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N_{(0, \infty)}(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.4), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

@@

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

@@

\section{Computation}\label{s:comp}
The MCMC for this model is fairly straightforward.
First, we impute values below the threshold.
Then, we update $\Theta$ using random walk MH or Gibbs sampling when appropriate.
Finally, we make spatial predictions using conditional multivariate normal results and the fact that the distribution of $Y_t(\bs) \mid \Theta, z_{tl}$ is the usual multivariate normal distribution with a \Matern spatial covariance structure.

We can use Gibbs sampling to update $Y_t(\bs)$ for censored observations that are below the threshold $T$.
After conditioning on $\alpha$, $z_t(\bs)$ and non-censored observations, $Y_t(\bs)$ has truncated normal full conditionals.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}(\mu(\bs), \Sigma)$.
After imputing the censored observations, we update the model parameters.
To update the model parameters, we use standard Gibbs updates for parameters when possible.
In the case Gibbs sampling is not possible, parameters are updated using a random-walk Metropolis Hastings algorithm.
See Appendices A.1 and A.2 for details regarding the MCMC.
The final step of the computation is to use Bayesian Kriging to generate a predictive distribution for $Y_t(\bs^*)$ at prediction location $(\bs^*)$.
This step is similar to the imputation for censored observations except that the full conditionals are no longer truncated at $T$.

@@

$M = 50$ data sets are generated for each design.
For each data set we fit the data using
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=1$ knots, $T=-\infty$
  \item $t$ marginal, $K=1$ knots, $T=q(0.80)$
  \item skew-$t$ marginal, $K=5$ knots, $T=-\infty$
  \item $t$ marginal, $K=5$ knots, $T=q(0.80)$
\end{enumerate}
where $q(0.80)$ is the 80th sample quantile of the data.
The design matrix $\bX$ includes an the intercept with a prior of $\beta \sim \text{N}(0, 10)$.
The spatial covariance parameters have priors $\log(\nu) \sim \text{N}(-1.2, 1)$, $\gamma \sim \text{U}(0, 1)$, $\log(\rho) \sim \text{N}(-2, 1)$.
The skewness parameter has prior $\alpha \sim \text{N}(0, 2)$.
The residual variance terms have priors $\sigma^2_t(\bs) \sim \text{IG}(0.1, 0.1)$.

@@

\section*{Appendix A.1: MCMC details}
The MCMC sampling for the model \ref{s:hier} is done using {\tt R} (http://www.r-project.org). Whenever possible, we select conjugate priors (see Appendix A.2); however, for some of the parameters, no conjugate prior distributions exist.
When no conjugate prior distribution exists, we use a random walk Metropolis Hastings update step.
In each Metropolis Hastings update, we tune the algorithm to give acceptance rates near 0.40.d

\subsection*{Spatial knot locations}
For each day, we update the spatial knot locations, $\bw_1, \ldots, \bw_K$, using a Metropolis Hastings block update.
First, we generate candidate knots locations using a random walk Gaussian candidate distribution
\begin{align*}

\end{align*}

\subsection*{Spatial covariance parameters}
We update the three spatial covariance parameters, $\log(\rho)$, $\log(\nu)$, $\gamma$, using a Metropolis Hastings block update step.
First, we generate a candidate using a random walk Gaussian candidate distribution 
\begin{align*}
	\log(\rho)^{(c)} \sim \text{N}(\log(\rho)^{(r - 1)}, s^2)
\end{align*}
where $\log(\rho)^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
Candidates are generated for $\log(\nu)$ and $\gamma$ in a similar fashion.
The acceptance ratio is 
\begin{align*}
	R = \left\{ \frac{ \prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(c)}, \nu^{(c)}, \gamma^{(c)}, \ldots] }{\prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(r-1)}, \nu^{(r-1)}, \gamma^{(r-1)}, \ldots] } \right\} \times \left\{ \frac{ p[\rho^{(c)}] }{ p[\rho^{(r - 1)] } } \right\} \times \left\{ \frac{ p[\nu^{(c)}] }{ p[\nu^{(r - 1)}] } \right\} \times \left\{ \frac{ p[ \gamma^{(c)} ] }{ p[\gamma^{(r - 1)} ] } \right\}
\end{align*}
where $l$ is the likelihood given in \ref{eq:hier}.
All three candidates are accepted with probability $\min\{R, 1\}$.

\section*{Appendix A.2: Posterior distributions}
\input{zpost.tex}
\input{betapost.tex}
\input{sigpost.tex}
\input{alphapost.tex}

\section*{Appendix A.3: Proof that $\lim_{h \rightarrow \infty} \pi(h) = 0$}
\input{proof_dist.tex}

\section*{Appendix A.4: Half-normal distribution}
Let $u = |z|$ where $Z \sim N(\mu, \sigma^2)$.
Specifically, we consider the case where $\mu = 0$. Then $U$ follows a half-normal distribution which we denote as $U \sim HN(0, 1)$, and the density is given by
\begin{align}
  f_U(u) = \frac{ \sqrt{2} }{ \sqrt{\pi \sigma^2} } \exp \left( - \frac{ u^2 }{ 2 \sigma^2 } \right) I(u > 0)
\end{align}
When $\mu = 0$, the half-normal distribution is also equivalent to a $N_{(0, \infty)}(0, \sigma^2)$ where $N_{(a, b)}(\mu, \sigma^2)$ represents a normal distribution with mean $\mu$ and standard deviation $\sigma$ that has been truncated below at $a$ and above at $b$.
